D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-06 ootavacuous.tex
Add example from P1916R0 and note per-execution sdep
Reported-by: David Goldblatt <davidtgoldblatt@gmail.com>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

Note that C++ implementations are permitted to evaluate to a more
general notion of OOTA values in cases involving unspecified or undefined
behavior, such as use of uninitialized objects.
However, these situations do not involve OOTA cycles, but rather user
errors and other issues that can lead to unfriendly optimizations that
in turn result in unpredictable output.
This paper focuses instead on OOTA cycles.

@@

@@

@@

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{a[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is undefined behavior, but the CPU neither knows
	nor cares, nor should it.}
Then line~7 can cause the final value of \co{r1} to be 42, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

>>>>

Note that C++ implementations are permitted to evaluate to a more
general notion of OOTA values in cases involving unspecified or undefined
behavior (UB), such as use of uninitialized objects.
However, these situations do not involve OOTA cycles, but rather user
errors and other issues that can lead to unfriendly optimizations that
in turn result in unpredictable output.
This paper focuses instead on OOTA cycles.

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and proposes a way to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.

@@

For example, in Listing~\ref{lst:Speculated Store and Non-Speculated Load},
suppose that the storage for \co{x} is located immediately before that
for \co{a[]}.
Suppose further that hardware speculation incorrectly guesses the value
of \co{i} to be -1, so that the speculated store of 42 on line~5 uses
the address of \co{x}.\footnote{
	Yes, \co{y[-1]} is UB, but the CPU neither knows
	nor cares, nor should it.}
Then line~7 can cause the final value of \co{r1} to be 42, when
by the rules of the C++ abstract machine the value must instead be
either zero or one.

@@

\section{Aside on Undefined Behavior}
\label{app:Aside on Undefined Behavior}

The combination of OOTA cycles and UB has proven especially
challenging~\cite{DavidGoldblatt2019NoElegantOOTAfix}.

\subsection{Alignment-Based Undefined Behavior}
\label{sec:Alignment-Based Undefined Behavior}

The example shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior (P1916R0)}
(taken from P1916R0) is especially instructive.

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   *(T*)r1 = *(T*)r1;
 8 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior (P1916R0)}
\label{lst:Alignment-Based Undefined Behavior (P1916R0)}
\end{listing}

Because the compiler often assumes that UB cannot happen, it might
look at line~7 and conclude that the value of \co{r1} necessarily
corresponded to an address aligned per the requirements of type \co{T}.
If that type's alignment was larger than four bytes, the compiler could
then replace the condition in line~5 with \co{true}, which would mean
that there was no semantic dependency from the load of \co{x} to the
store of \co{y}.

Except that the developer might have knowingly carried out a misaligned
access, for example, to test an alignment trap handler or because the
system on which this code was to run was known to gracefully handle
misaligned accesses.

\begin{listing}[tbp]
\begin{verbatim}
 1 std::atomic<int> y;
 2 template <typename T>
 3 void t1() {
 4   long long r1 = x.load(std::memory_order_relaxed);
 5   if (r1 % 4 == 0)
 6     y.store(1, std::memory_order_relaxed);
 7   __asm__ __volatile__("": : :"mightnotreturn");
 8   *(T*)r1 = *(T*)r1;
 9 }
\end{verbatim}
\caption{Alignment-Based Undefined Behavior Fix}
\label{lst:Alignment-Based Undefined Behavior Fix}
\end{listing}

However, in that case, the developer is stepping outside the bounds
of the standard.
It would be good to have some way for the developer to get the job
done without confusing the compiler, thus separating OOTA-cycle and
UB concerns.
One approach is to have a \co{mightnotreturn} clobber similar to the
\co{memory} clobber for GCC asms.
Then this example could be rewritten as shown in
Listing~\ref{lst:Alignment-Based Undefined Behavior Fix},
with the asm on line~7 preventing the compiler from assuming that line~8
is guaranteed to be executed, thus preventing the compiler from
applying any lessons that it might learn from line~8 to line~5.

% Clean up <200b> in vim: ":%s/\%u200b//g"

This \co{mightnotreturn} clobber is merely an example.
There might well be better ways to prevent UB from propagating backwards.
Other options being considered include a call to an empty function that
is invisible to the compiler and a GCC goto asm that the compiler cannot
prove falls through.

\subsection{Unreachability-Based Undefined Behavior}
\label{sec:Unreachability-Based Undefined Behavior}

This section attempts to apply the lessons
of Appendix~\ref{sec:Unreachability-Based Undefined Behavior} to
Goldblatt's unreachability-UB example.  Inserting the asm at line~12
as shown in Listing~\ref{lst:Alignment-Based Undefined Behavior Fix}
avoids confusing the compiler into assuming that the value of \co{local}
must necessarily be zero.

\begin{listing}[tbp]
\begin{verbatim}
 1 // UB if this function is ever called.
 2 [[noreturn]]
 3 inline void unreachable() {
 4   return;
 5 }
 6 void t1() {
 7   r1 = x.load(std::memory_order_relaxed);
 8   int local = (r1 & 1);
 9   if (local == 0)
10     y.store(1, std::memory_order_relaxed);
11   if (local != 0) {
12     __asm__ __volatile__("": : :"mightnotreturn");
13     // If we get here, UB.
14     unreachable();
15   }
16 }
\end{verbatim}
\caption{Unreachability-Based Undefined Behavior Fix}
\label{lst:Unreachability-Based Undefined Behavior Fix}
\end{listing}

By the same reasoning, Goldblatt's \co{misbhave} example that
also involves unreachability can also be addressed through insertion
of the asm.
